---
title: "Typical Filipino"
output: 
  html_document:
    keep_md: true
    theme: spacelab
    highlight: monochrome
    css: style.css
---

# Introduction
Filipinos are cheerful, resourceful and practical people. It is in our nature 
to make the best out a of any situation, no matter how hard it is. We get this 
from our strong family values and religious beliefs. However, our ideals are 
not always represented in  our actions especially when it comes to our 
financial priorities. I would hear Filipinos complaining about low salaries but
are spending so much on things they do not need. 

## Objectives
In this article, I would like to find out if the spending habits and financial 
priorities change as their income increases of a typical Filipino family.

## Dataset Source
This dataset is from The Philippine Statistics Authority who conducts the 
Family Income and Expediture Survey (FEIS) every 3 years nationwide. This is 
from the 2015 most recent survey. The raw data was cleaned by Francis Paul 
Flores from his [kaggle.com/grosvenpaul](https://www.kaggle.com/grosvenpaul/family-income-and-expenditure/downloads/Family%20Income%20and%20Expenditure.csv).

# Getting a Feel of the Data
The Dataset contains 41,544 observations of Filipino Households from every 
Region of the country. It is comprised of 60 variables describing each family 
on their income, family description and expenditure.

```{r Load_data, message=FALSE, warning=FALSE}
library(readr)

fies <- read_csv("fies.csv")

names(fies)
```

```{r renameCols}
# Rename Lables for easier reference.
names(fies) <- c("income", "region", "expense", "source", 
                 "agri_house", "exp_bread", "exp_rice", "exp_meat", 
                 "exp_seafood", "exp_fruit", "exp_veg", "exp_resto_hotel", 
                 "exp_alcoh", "exp_taba", "exp_clothe", "exp_house_water", 
                 "exp_rent", "exp_med", "exp_trans", "exp_comms", 
                 "exp_edu", "exp_misc", "exp_spec", "exp_farm",
                 "inc_entrep", "head_gender", "head_age", "head_stat", 
                 "head_educ", "head_job_bus", "head_occup", "head_workclass", 
                 "family_t", "family_n", "baby_n", "kid_n", 
                 "employed_n", "house_t", "roof_t", "wall_t", 
                 "house_area", "house_age", "bed_n", "house_tenure",
                 "toilet", "electric", "water_t", "tv_n", 
                 "DVD_n", "sterio_n", "ref_n", "wash_n", 
                 "aircon_n", "car_n", "tel_n", "cell_n",
                 "pc_n", "stove_n", "mboat_n", "mbike_n")

names(fies)
```

There are 60 variables in this dataset and are not code-friendly. First thing 
to do is rename each column key. A detailed desciption of each column is 
documented on a separate txt file for your reference.

## Income Distribution

```{r incomeDistribution, message=FALSE, warning=FALSE}
library(ggplot2)

ggplot(fies, aes(income)) +
  geom_histogram()

```

It's no surprise that income per household is skewed to the left, where there
is more poor families and few rich people with dramatic outlying incomes.

```{r incomeSummary}
summary(fies$income)
```

Doing a quick summary of the income, we see that the average income of each
household is 247,556 Php annually. But this is skewed because of the maximum
income of 11 mill Php of the super rich.

## Averaging Everything is just Lazy

```{r incomeLog10, echo=FALSE, message=FALSE, warning=FALSE}
library(scales)
set.seed(222)
fies_sam <- fies[sample(nrow(fies), 10000), ]
ggplot(fies_sam, aes(income)) +
  geom_histogram(col = I("white")) +
  scale_x_continuous(
      trans = log10_trans(),
      breaks = c(10000, 25000, 70000,
                 150000, 247556, 400000,
                 1000000, 2000000)) +
  geom_vline(data = fies, 
             aes(xintercept = mean(fies$income)), 
             color = I("red"))
```

We transform the histogram to adjust the power of magnitude of each income and
as a result we can see a beautiful normal distribution. To understand why we
use log10, imagine that we made the x axis points closer together so that we
can see see the distribution more clearly.

The problem with relying on averages is that it is heavily influenced by 
outliers. Php 247,556 in the graph is far from the center, which makes it a
poor representative of the distribution.

The Rich are really rich at 11 mill Php  and the Poor are really poor with 
only 11,285 Php of annual income.

In addition, this average is of the whole population. We can't do that because,
for some people, 247k is small while other it is very big. So, we facet the 
graph according to different groups or work classes.

## Grouped by Working Class according to Survey

```{r groupByWorkclass, message=FALSE, warning=FALSE}
fies$head_occup <- factor(fies$head_occup)
fies$head_workclass <- factor(fies$head_workclass)

ggplot(fies, aes(income)) +
  geom_histogram() +
  scale_x_continuous(
                     trans = log10_trans(), 
                     breaks = c(10000, 25000, 70000,
                                150000, 400000, 
                                1000000, 2000000)) +
  facet_wrap(~ head_workclass)
```

In this graph, we use the work class determined by the survey. But this is not 
really telling us anything especially that the categories are vague and are 
not familiar. The needs a little bit of cleaning.

## We make our own Groups

```{r groupByCustomClass, message=FALSE, warning=FALSE}
library(stringr)
library(dplyr)

pro_class <- paste(c("managers", "medical", "dentist", "lawyer",
                     "architect", "supervisor", "trade", "education",
                     "chemist", "professionals", "scientist",
                     "justice", "Mechanical engineers", "director",
                     "accountant", "Chemical engineers", 
                     "Electrical Engineers", "Industrial engineers",
                     "designer", "singers", "television", "dancer",
                     "fashion", "marketing", "Civil engineers", "air",
                     "Computer programmers", "nurse", "Librarians",
                     "science", "therapist", "opticians", "Veter",
                     "pilots", "estate agents", "plant operators", 
                     "legislative"),
                   collapse = "|")

farm_class <- paste(c("farm", "fish", "growers", "gather", "duck",
                      "cultivat", "plant", "hunter", "village",
                      "practitioners", "healers", "diary", "animal"), 
                    collapse = "|")

work_class <- paste(c("workers", "carpenters", "welder", "helper", "clean",
                      "launder", "labor", "driver", "freight", "mechanics",
                      "mason", "porter", "waiter", "cook", "foremen",
                      "caretaker", "lineman", "varnish", "conductor",
                      "salesperson", "bookkeeper", "inspector", "undertaker", 
                      "loggers", "wood", "roofers", "cutter", "electricians",
                      "assembler", "builder", "metal",  "tanners", "garbage",
                      "repair", "prepare", "rigger", "vendors", "valuers",
                      "setters", "guides", "tasters", "potters", "preservers",
                      "textile", "fitters", "valets", "blasters", 
                      "humanitarian", "staff officers"), 
                    collapse = "|")

off_class <- paste(c("pawn", "buy", "baker", "maker", "tailor", "assistant",
                     "clerk", "engineering technicians", "operators",
                     "artists", "managers/managing proprietors", "insurance",
                     "educ", "principals", "secretaries", "general", 
                     "pharmacists", "commercial", "communications engineers",
                     "draftsmen", "instructors", "travel consultants",
                     "enlisted personnel"), 
                    collapse = "|")

fies$head_class <- with(fies,
                        ifelse(str_detect(head_workclass, "government"), 
                               "gov",
                        ifelse(grepl(work_class, head_occup, ignore.case = T),
                               "work",
                        ifelse(grepl(off_class, head_occup, ignore.case = T),
                               "off",
                        ifelse(grepl(farm_class, head_occup, ignore.case = T),
                               "farm",
                        ifelse(grepl(pro_class, head_occup, ignore.case = T),
                               "pro",
                               NA))))))
```

## We got to Catch them All

```{r Reveiw_class, warning=FALSE, paged.print=FALSE}

fies.occup <- fies %>%
  filter(is.na(fies$head_class)) %>%
  group_by(head_occup) %>%
  summarise(n= n(),
            income_mean = mean(income),
            income_median = median(income)) %>%
  arrange(desc(n))

fies.occup
```

I decided to evaluate each occupation of the head of the family and manually 
group each household according to the following working class:

- *Professionals.* Jobs that usually requires higher education.
- *Office* Usually found in shops and offices of goods.
- *Government.* Jobs in government.
- *Workers.* Jobs that usually pays low and does not require a degree.
- *Farmers.* Jobs that pertains to rural work. Agriculture, Fishing, etc.

I am basing the categories based on my own classifications of how I see the
working class and their overall mean income. I am ignoring the households
that did not specify the occupation of the head of the family.

```{r summaryCustomClass }
tapply(fies$income, fies$head_class, summary)
```

## A Millioniare Waiter

I review each working class and see what I can further refine. My goal is to
segment as much as I can to emphasize the division among the working classes
and at the same time being fair with my evaluation of the different jobs.

```{r reveiwWorkerClass, message=FALSE, warning=FALSE}
fies.workers <- fies %>%
  filter(fies$head_class == "work") %>%
  select(head_occup, income, inc_entrep) %>%
  arrange(desc(income), head_occup)

fies.workers
```

```{r findingWaiterOutlier, message=FALSE, warning=FALSE}
subset(fies.workers, income == "5652261")
```

I then checked the worker class and noticed some interesting outliers in the
group. The one that stood out was a Waiter job who is earning Php5,652,261
annually. This does not seem right to me so I took a deeper look and found out
that income is the sum of their income from their jobs and their business.

Subtracting the income from business from the total income, the Waiter was 
actually earning Php544,810 annually from his job and Php5,107,451 from his 
business. For Analysis, we would like to see the income from their employment.

## We just want the Income from their Jobs

```{r addIncomeFromWork}
fies$inc_work <- with(fies, income - inc_entrep)

# Check if there are no negative income from work.
head(table(sort(fies$inc_work)))
```

```{r summaryOfCustomGroups}
tapply(fies$inc_work, fies$head_class, summary)
```

## Reveiwing the Professionals

```{r reviewProfessionals, message=FALSE, warning=FALSE}
fies.pro <- fies %>%
  filter(fies$head_class == "pro") %>%
  group_by(head_occup) %>%
  summarise(income_mean = mean(inc_work),
            income_median = median(inc_work),
            n= n()) %>%
  arrange(desc(income_mean))

fies.pro
```

## Reviewing the Office Employees

```{r reviewOffice, message=FALSE, warning=FALSE}
fies.office <- fies %>%
  filter(fies$head_class == "off") %>%
  group_by(head_occup) %>%
  summarise(income_mean = mean(inc_work),
            income_median = median(inc_work),
            n= n()) %>%
  arrange(desc(income_mean))

fies.office
```

## Reviewing the Farmers

```{r reviewFarmer, message=FALSE, warning=FALSE}
fies.farm <- fies %>%
  filter(fies$head_class == "farm") %>%
  group_by(head_occup) %>%
  summarise(income_mean = mean(inc_work),
            income_median = median(inc_work),
            n= n()) %>%
  arrange(desc(income_mean))

fies.farm
```

## See what happens when you just Average

```{r countWorkclass}
fies$head_class <- factor(fies$head_class, 
                          levels = c("pro", "gov", "off", "farm", "work"),
                          ordered = T)

ggplot(subset(fies, !is.na(head_class)), aes(head_class)) +
  geom_bar()
```

In our dataset, we generally have more workers than any of the other classes.
Professionals are the minority of the group.

```{r incomeDistributionByCustomClass, message=FALSE, warning=FALSE}
ggplot(subset(fies, income != 0 & 
                !is.na(head_class)), 
       aes(income)) +
  geom_histogram(col=I("white")) +
  scale_x_continuous(trans = log10_trans(), 
                     breaks = c(10000, 25000, 70000,
                                247556, 
                                500000, 2000000),
                     limits = c(2000, 2500000)) +
  facet_wrap( ~ head_class, ncol = 2)  +
  geom_vline(data = subset(fies, !is.na(head_class)),
             aes(xintercept = mean(fies$income)),
             color = I("red"))
```

We go back to the overall income distribution but this time using our custom
work class. We add the average income Php247k and we see that some groups
are below the average line and other groups are above the average line.

## Income from Salaries

```{r incomeFromSalaries, message=FALSE, warning=FALSE}
# We remove the households that does not specify occupation.
ggplot(subset(fies, inc_work != 0 & 
                !is.na(head_class)), 
       aes(inc_work)) +
  geom_histogram(col=I("white"), fill = I("#00CCFF")) +
  scale_x_continuous(trans = log10_trans(), 
                     breaks = c(10000, 25000, 70000,
                                247556, 
                                500000, 2000000),
                     limits = c(2000, 2500000)) +
  scale_y_continuous(breaks = c(0, 500, 1000, 1500,
                               2000, 2500),
                     limits = c(0, 2500)) +
  facet_wrap( ~ head_class, ncol = 2)  +
  geom_vline(data = subset(fies, !is.na(head_class)),
             aes(xintercept = mean(fies$income)),
             color = I("red"))
```

If people relied from their salaries, the picture is the same. Farmers, Office
and Workers are below the average line while Gov Employees and Professionals
are above the average line.

## Income from Business

```{r expenseByCustomClass, message=FALSE, warning=FALSE}
ggplot(subset(fies, inc_entrep != 0 & 
                !is.na(head_class)), 
       aes(inc_entrep)) +
  geom_histogram(col=I("white"), fill = I("#82AF4C")) +
  scale_x_continuous(trans = log10_trans(), 
                     breaks = c(10000, 25000, 70000,
                                247556, 
                                500000, 2000000),
                     limits = c(2000, 2500000)) +
  scale_y_continuous(breaks = c(0, 500, 1000, 1500,
                               2000, 2500),
                     limits = c(0, 2500)) +
  facet_wrap( ~ head_class, ncol = 2)  +
  geom_vline(data = subset(fies, !is.na(head_class)),
             aes(xintercept = mean(fies$income)),
             color = I("red"))
```

If people relied solely on income from their business, all of them are below
the Average line of the overall income population. This shows that majority of
Filipinos have higher income from Salaries than from business.

## View from BoxPlots

```{r boxplotsIncomeSalaries}
ggplot(subset(fies, !is.na(head_class)), aes(head_class, inc_work)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, quantile(fies$inc_work, .98)))
```

```{r boxplotsIncomeBusiness}
ggplot(subset(fies, !is.na(head_class)), aes(head_class, inc_entrep)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, quantile(fies$inc_entrep, .92)))
```

Both graphs shows that professionals are heavily dependent on salaries from
their professions as their main source of income. While most groups would treat
treat their business as a secondary source of income, office employees and
farmers would seem to have more profitable business than the rest of the
groups.

But this is not clear because we are mixing households that rely on salaries
and business alone with people who are soley relient on either their salaries 
or their business.

## Comparing Source of Income in One Graph

```{r transformWideToLong}
library(tidyr)

fies$id <- seq_len(nrow(fies))

fies.n <- fies %>%
  filter(!is.na(fies$head_class)) %>%
  select(id, head_class, income, inc_entrep, inc_work, expense) %>%
  gather(inc_type, inc, c(inc_work, inc_entrep)) %>%
  gather(cashflow, cash, c(income, expense)) %>%
  arrange(id)

fies.n
```

We would like to compare work income from business income to find out the
relationship. We transform the data from wide format to long format so that
we can graph the data in a frequency polygon.

```{r salaryVsBusiness, message=FALSE, warning=FALSE}
ggplot(fies.n, aes(inc, color = inc_type)) +
  geom_freqpoly(alpha =.5) +
  scale_x_continuous(trans = log10_trans(), 
                     breaks = c(10000, 25000, 70000,
                                247556, 
                                500000, 2000000),
                     limits = c(2000, 2500000)) +
  facet_wrap(~ head_class, ncol = 2)
```

Placing the graphs on a frequency polygon, we see that farmers have both their
salaries from business income at almost similar distribution with their salary
income. Office Employees have income from Salaries higher than their business
income. While workers would have a distribution that have a higher income from
Salaries and income from business more spread out. Government employees and
Professionals does not have a clear relationship between income from salaries
business.

## Returning to our Question

```{r}
fies.w <- fies %>%
  filter(!is.na(head_class)) %>%
  select(id, head_class, income, expense)
```


```{r}
ggplot(fies.w, aes(income, expense)) +
  geom_point(alpha = .2) +
  scale_x_continuous(limits = c(0, quantile(fies$income, .98))) +
  scale_y_continuous(limits = c(0, quantile(fies$expense, .98)))
```

We want to relate expense with their income and if we graph the relationship,
we see that as income increase, so does their expenses, which is obvious.

```{r}
ggplot(fies.w, aes(income, expense, color = head_class)) +
  geom_point(alpha = .2) +
  scale_x_continuous(limits = c(0, quantile(fies$income, .98))) +
  scale_y_continuous(limits = c(0, quantile(fies$expense, .98))) +
  scale_color_brewer(type = "div")
```

If we look at the same graph but this time color-coded according to our custom
class, we see better which types of proffesions are having higher incomes. But
this is not really clear.

```{r}
ggplot(fies, aes(income, expense, color = head_class)) +
  geom_point(alpha = .2) +
  scale_x_continuous(limits = c(0, quantile(fies$income, .98))) +
  scale_y_continuous(limits = c(0, quantile(fies$expense, .98))) +
  facet_wrap(~head_class)
```
 
 Splitting the graphs in a facet, it seems that all classes are resembling the
 same pattern where the more income they have the higher they would spend. This
 does not tell us anything new. And I get to thinking, what are they actually
 spending on? To answer that, we look at our available variables and see.
 
```{r}
names(fies)
head(fies)
```

```{r}
fies.w.exp <- fies %>%
  filter(!is.na(head_class)) %>%
  mutate(exp_food = exp_bread + exp_rice + exp_meat +
           exp_seafood + exp_fruit + exp_veg,
         exp_lux = exp_resto_hotel + exp_spec + exp_clothe,
         exp_vice = exp_alcoh + exp_taba,
         exp_util = exp_house_water + exp_rent + exp_comms + exp_trans +
           exp_misc) %>%
  gather(cashflow, cash, c(exp_food, exp_lux, exp_vice, exp_util,
                           exp_med, exp_edu, exp_farm)) %>%
  select(id, head_class, income, expense, cashflow, cash) %>%
  arrange(id)

names(fies.w.exp)
fies.w.exp$exp_type <- factor(fies.w.exp$exp_type)
levels(fies.w.exp$exp_type)
summary(fies.w.exp)
fies.w.exp
```

```{r}
ggplot(fies.w.exp, aes(x = income, color = cashflow, fill = cashflow)) +
  geom_density() +
  scale_x_continuous(trans = log10_trans())


```

```{r}
ggplot(fies.w.exp, aes(income, cash, color = head_class)) +
  geom_point(alpha = .2) +
  scale_x_continuous(trans = log10_trans()) +
  scale_y_continuous(trans = log10_trans(),
    limits = c(10000, quantile(fies.w.exp$cash, .98))) +
  facet_wrap(~cashflow)

```

